A Transactional Memory System for Parallel Python

### 2014-06-08 talk/dls2014/paper/paper.tex
my last changes

Since some years, the popularity of dynamic languages is on the rise.
A common trait of many of these language implementations is the use of
a single global interpreter lock (GIL) to synchronise the interpreter
in a multithreading scenario. Since this lock effectively serialises
the execution, applications can not make use of the increasing parallelism in
current hardware.

In this paper, we present a software transactional memory (STM) system
specifically designed for the purpose of replacing the GIL in dynamic
language interpreters. The key idea is a close integration of STM with
the garbage collector to approach atomicity, synchronisation, and
memory management in a unified way.  As a user abstraction for a
scalable synchronisation mechanism, we implement atomic blocks by
exposing part of the STM system to applications.  The described STM
uses a combination of features present in current CPUs, so that the
parallelisation benefits outweigh the system's overhead already on 2
threads when comparing to a GIL supported Python interpreter.

@@

Unfortunately the GIL plays a central role in current
implementations. Even recent just-in-time compilers (JIT) for these
languages (e.g., PyPy~\cite{cfbolz09}, V8~\cite{kevin10},
IonMonkey~\cite{ionmonkey}) are based on the GIL; Jython (an
implementation of Python that is based on the JVM) is a notable
exception.

The standard model employed in these systems is fairly
straightforward: there can be multiple threads, but each acquires the
GIL, then performs one (or several) instruction, and subsequently releases the
lock.  While this setup limits the benefits due to parallelism,
the use of a GIL also provides
some crucial guarantees. Since this lock is always acquired while
executing bytecode instructions and may be released only in-between
such instructions, the GIL ensures perfect isolation between multiple
threads for a series of instructions, which are executed atomically.
In addition, the GIL provides the application with a sequential
consistency model~\cite{lamport79}.

A global interpreter lock assures the atomic execution of individual
instructions but offers the programmer no help in managing concurrent
data structures. Many parallel programs must maintain invariants
involving multiple objects (or even multiple fields of a single
object).  So to enable Python programs for parallel execution, we need
a way to identify which blocks of data must be executed atomically and
hardware or software support to effect this atomic execution.
This is traditionally done with locks.

\emph{Atomic blocks}~\cite{tim03,tim05} would allow dynamic languages to
indicate the units of computation that must be executed atomically to
ensure correct concurrent execution.  Atomic blocks map nicely to
transactional memory, which in turn provides a convenient mechanism to replace
the GIL.  Using transactions to enclose multiple bytecode
instructions, we can get the very same semantics as provided by the
GIL while possibly executing several transactions in parallel. And by
exposing these interpreter-level transactions to the application, we
avoid the well-known problems caused by locks as applications use
atomic blocks instead of explicit acquire/release operations on
locks~\cite{christopher10,victor11,shan08}.

@@

This paper makes the following contributions:
%\vspace{-\topsep}               % XXX:HACK:TODO
\begin{itemize}
\item We introduce a new STM system that performs well for dynamic
  language interpreters.
\item We integrate the system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}

@@

TM also allows us to attack the issue of application-level
synchronisation and atomicity. We introduce atomic blocks to the
Python language by mapping them directly to the underlying
transactions -- providing the application with the guarantee of
atomicity for the enclosed instructions.

@@

Clearly, if we are in a shared-memory setting, this model is extremely
wasteful: at any given time, most objects are identical in all, or
most, segments.  They differ only in segments that have modified but
not yet committed these objects.  So we make use of the Memory
Management Unit (MMU) of the CPU.  The MMU maps pages of virtual to
physical memory.  By changing this page mapping, at the coarse unit of
one page of memory (4096 bytes), we can share memory that would
otherwise exist as identical copies in all segments.

@@

\subsection{Memory Segmentation on the x86}

@@

What we described so far has no measurable performance overhead by
itself, but it is missing conflict detection to be considered a
full-fledged STM system.  This requires adding \emph{barriers} to the
program to register the objects in the read or write set before
reading or writing to them\footnote{Barriers added by an automatic and
local program transformation~\cite{felber07}}.  Furthermore, we still
need to describe the commit protocol.

@@

\lstinline!stm_start_transaction()!  starts a transaction in the
current thread. Internally, it uses \lstinline!setjmp()! to remember
where the transaction started in case of an abort.
\lstinline!stm_commit_transaction()! tries to commit the current
transaction, aborting it otherwise. Similarly,
\lstinline!stm_become_inevitable()!  tries to make the current
transaction inevitable, possibly aborting the current one if not
possible.
 \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on the object given as the argument, and
\lstinline!stm_allocate()!  allocates a new object with the specified
size.  \lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
push and pop objects onto the shadow stack\footnote{A stack for pointers
  to GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.~\cite{fergus02}} of the current thread.  References to objects
have to be saved using this stack locally around calls that may cause a GC
cycle to happen, and also while there is no transaction
running. Otherwise these references may not be valid any more because
the object moved or was seen as garbage. In this simplified Application
Programming Interface (API), only \lstinline!stm_allocate()!,
\lstinline!stm_become_inevitable()! and
\lstinline!stm_commit_transaction()!  require saving object
references.

@@

It may not be possible to immediately break the transaction, as we
could be executing an atomic block. Atomic blocks are currently
implemented as guaranteeing that the whole block's execution takes
place in a single transaction. Another issue is the support of
irreversible operations, like calls to external functions or general
input~/ output. If we are allowed to, we simply commit the transaction
before and start a new one after these operations. Within an atomic
block, however, we turn the transaction inevitable so that the
transaction cannot abort and strong isolation is guaranteed.

@@

On startup, we reserve a big range of virtual memory with a call to
\lstinline!mmap()! and partition this space into $N+1$ segments. We
want to run $N$ threads in parallel while segment~0 is designated as
the \emph{sharing-segment} that is never assigned to a thread. The
sharing-segment is essentially read-only, except during commit.

@@

Note, these areas are at offsets inside the segments known at
compile time. This makes some checks very efficient, e.g.\ checking
if an object resides in the nursery only requires comparing its $SO$
to the boundaries of the nursery.

@@

Furthermore, we optimistically re-share pages during major collections.
This is another example where the STM-GC cooperation makes sense. We
regard private pages as additional memory that may be collected.
Pages should usually stay private for some time, but at some regular
intervals, we still want to try to re-share pages in case they actually
stay unmodified in the future. Major collections are allowed to do
that because all threads are synchronised, and we can re-share all
private pages that do not currently contain an object in some
transaction's  write set.

@@

The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.  For
each object, this code is called at most once per minor collection.
First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!). Any reference we write to this object will
be kept alive and up-to-date during minor collections.

@@

We evaluate our system in a Python interpreter called
PyPy~\cite{pypy}.
\begin{description}
\item[PyPy] (version 2.3) is an implementation of an
  interpreter for the Python language. It has a special focus on speed
  and provides a just-in-time (JIT) compiler to speed up applications
  running on top of it. We compare between normal PyPy
  using a GIL and a PyPy with STM, as well as to CPython.
\item[CPython] (version 2.7.6) is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\end{description}

@@

For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
and normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread ($12.3\%$ for all 4 threads together). Every thread
adds some overhead because during a commit, there is one more thread
which has to reach a safe point. Additionally, conflict detection
during commit needs to check in all concurrently running transactions.

@@

\subsection{Non-JIT Benchmarks}
First we run our benchmarks on three different interpreters: CPython
(GIL), PyPy with STM, and PyPy with the GIL (all without the JIT). The
results are shown in Figure~\ref{fig:performance-nojit}.

@@

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones ($avg=2.6\times$ speedup) and a little less for
the others ($avg=2.0\times$ speedup). The reason for that is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks to synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

@@

Since PyPy comes with a JIT to make its overhead compared to CPython
go away, we will now look at how well STM works together with it.

@@

We would like to regard enabling the JIT as a simple performance
enhancement, but that is not what happens in reality. First, since the
JIT~\cite{cfbolz09} bases its decision of what to compile on runtime
profiling, running in multiple threads, it may compile different
things in each run because of the non-deterministic thread-scheduling
of the operating system (OS). Second, it is able to remove some
allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.
For these reasons, the following results have to be taken with a grain
of salt.

@@

There are multiple attempts at removing the limits of HTM by means of
\emph{virtualising} HTM~\cite{rajwar05,chung06}. Indeed this may also
be the future direction of our research. So far, some of them depend
on additional hardware features. Or in the case of~\cite{chung06},
they have the assumption that most transactions fit the HTM capacity
and that virtualisation is only a fallback. Our use case, however,
features arbitrarily long transactions, which is why this assumption
first has to be verified in our case.

@@

% STM needs good performance:
The solution presented here is based on a software TM system.  As the
STM is a vital part of the interpreter, good performance is
essential. We present an interesting way to combine the existing
memory segmentation and virtual memory features of current CPUs to
lower the overhead of STM. This software implementation reduces
dramatically the cost of transactional memory and at the same time
provides an opportunity to integrate the STM system with garbage
collection.  This implementation leverages the processor's memory
segmentation hardware, which has lately played only a minor role in
our operating systems.  Unfortunately, processor architects seem to
undervalue this hardware resource -- in the evolution from the x86 to
the x86{-}64 architecture, 4 out of 6 segment registers have been
removed. This development potentially threatens the STM strategy
outlined here and we want to remind the architecture community that
this unit can support interesting implementations -- a viewpoint that
has also been expressed by other researchers (see also
\cite{bennet10}).

% results, outlook
The early results presented here are very encouraging. STM as a simple
GIL replacement yields an average speedup of $2.6\times$ for
embarrassingly parallel workloads on 4 threads.  When also used for
synchronisation in the form of atomic blocks, the average speedup
still reaches $2.0\times$.

To obtain an overall performance that is competitive with the
best-performing Python systems (Jython, PyPy), integration of the
STM-based approach with a JIT compiler is necessary. Once this
integration matures, the approach outlined here serves not only as a
simple GIL replacement but also provides a way forward towards
parallelising Python.

@@

\acks
We would like to thank the following people for their valuable
inputs and the many fruitful discussions: Carl F. Bolz, Michael Faes,
Pascal Felber, Maciej Fijalkowski, Patrick Marlier, and Aristeidis Mastoras.

>>>>

Since some years, the popularity of dynamic languages is on the rise.
A common trait of many of these language's implementations is the use of
a single global interpreter lock (GIL) to synchronise the interpreter
in a multithreading scenario. Since this lock effectively serialises
the execution, applications can not make use of the increasing parallelism in
current hardware.

In this paper, we present a software transactional memory (STM) system
specifically designed for the purpose of replacing the GIL in dynamic
language interpreters while keeping the semantics. The key idea is a
close integration of STM with the garbage collector to approach
atomicity, synchronisation, and memory management in a unified
way. Additionally, we implement atomic blocks as a scalable
synchronisation mechanism by exposing part of this STM system to
applications.

The described system uses a combination of features present in current
CPUs, so that the parallelisation benefits outweigh the system's
overhead already on 2 threads when comparing two Python interpreters
-- one with STM and one with a GIL.

@@

Unfortunately the GIL plays a central role in current
implementations. Even recent efforts to address the performance problems
of dynamic languages with just-in-time compilers (JIT)
(e.g., PyPy~\cite{cfbolz09}, V8~\cite{kevin10},
IonMonkey~\cite{ionmonkey}) keep the GIL; Jython (an
implementation of Python that is based on the JVM) is a notable
exception.

The standard model employed in these systems is fairly
straightforward: there can be multiple threads, but each acquires the
GIL, then performs one (or several) instruction, and subsequently releases the
lock.  While this setup limits the benefits due to parallelism,
the use of a GIL also provides
some crucial guarantees. Since this lock is always acquired while
executing bytecode instructions and may be released only in-between
such instructions, the GIL ensures perfect isolation between multiple
threads for a series of instructions -- executing them atomically.
In addition, the GIL provides the application with a sequential
consistency model~\cite{lamport79}.

A global interpreter lock assures the atomic execution of individual
instructions but offers the programmer no help in managing concurrent
data structures. Many parallel programs must maintain invariants
involving multiple objects (or multiple fields of a single
object).  So to enable Python programs for parallel execution, we need
a way to identify which blocks of data must be executed atomically and
hardware or software support to effect this atomic execution.
This is traditionally done with locks.

\emph{Atomic blocks}~\cite{tim03,tim05} allow languages to
indicate the units of computation that must be executed atomically to
ensure correct concurrent execution.  Atomic blocks map nicely to
transactional memory, which in turn provides a convenient mechanism to replace
the GIL.  Using transactions to enclose multiple bytecode
instructions, we can get the very same semantics as provided by the
GIL while possibly executing several transactions in parallel. And by
exposing these interpreter-level transactions to the application, we
avoid the well-known problems caused by locks as applications use
atomic blocks instead of explicit acquire/release operations on
locks~\cite{christopher10,victor11,shan08}.

@@

This paper makes the following contributions:
%\vspace{-\topsep}               % XXX:HACK:TODO
\begin{itemize}
\item We introduce a new STM system that performs well for dynamic
  language interpreters.
\vspace{3mm}               % XXX:HACK:TODO
\item We integrate the system closely with a garbage collector
  (GC) to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated.
\end{itemize}

@@

TM also allows us to attack the issue of application-level
synchronisation and atomicity. We introduce atomic blocks to the
Python language by mapping them directly to the underlying
transactions -- providing the application with the guarantees of
atomicity and isolation for the enclosed instructions.

@@

Clearly, if we are in a shared-memory setting, this model is extremely
wasteful: at any given time, most objects are identical in all, or
most, segments.  They differ only in segments that have modified but
not yet committed these objects.  So we make use of the Memory
Management Unit (MMU) of the CPU.  The MMU maps pages of virtual to
physical memory.  By changing this page mapping at the coarse unit of
one page of memory (4096 bytes), we can share memory that would
otherwise exist as identical copies in all segments.

@@

\subsection{Memory Segmentation on x86 CPUs}

@@

What we described so far has no measurable performance overhead by
itself, but it is missing conflict detection to be considered a
full-fledged STM system.  This requires adding \emph{barriers} to the
program to register the objects in the read or write set before
reading or writing to them\footnote{Barriers added by an automatic, conservative, and
local program transformation~\cite{felber07}}.  Furthermore, we still
need to describe the commit protocol.

@@

\lstinline!stm_start_transaction()!  starts a transaction in the
current thread. Internally, it uses \lstinline!setjmp()! to remember
where the transaction started in case of an abort.
\lstinline!stm_commit_transaction()! tries to commit the current
transaction, aborting it otherwise. Similarly,
\lstinline!stm_become_inevitable()!  tries to make the current
transaction inevitable, waiting or aborting the current one if there
is already an inevitable transaction.
 \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on the object given as the argument, and
\lstinline!stm_allocate()!  allocates a new object with the specified
size.

\lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
push and pop objects onto the shadow stack\footnote{A stack for pointers
  to GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.~\cite{fergus02}} of the current thread.  References to objects
have to be saved using this stack locally, and conservatively around calls that may cause a GC
cycle to happen, and also while there is no transaction
running. Otherwise these references may not be valid any more because
the object moved or was discarded as garbage. In this simplified Application
Programming Interface (API), only \lstinline!stm_allocate()!,
\lstinline!stm_become_inevitable()! and
\lstinline!stm_commit_transaction()!  require saving object
references.

@@

It may not be possible to immediately break the transaction, as we
could be executing an atomic block. Atomic blocks are currently
implemented as guaranteeing that the whole block's execution takes
place in a single transaction.

Another issue is the support of
irreversible operations, like calls to external functions or general
input~/ output. If we are allowed to, we commit the transaction
before and start a new one after these operations\footnote{The language
specification needs to specify if this is allowed.}. Within an atomic
block, however, the transaction is unbreakable, and we need to turn
the transaction inevitable so that it
cannot abort and strong isolation is guaranteed.

@@

On startup, we reserve a big range of virtual memory with a call to
\lstinline!mmap()! and partition this space into $N+1$ segments.
We want to run $N$ threads in parallel while segment~0 is designated as
the \emph{sharing-segment} that is never assigned to a thread. The
sharing-segment is essentially read-only (except during commit) and
holds all pages shared between the other segments.

@@

Note, since the above configuration is specified at compile time,
all these areas are at offsets inside the segments known to the
compiler. This makes some checks very efficient, e.g.\ checking
if an object resides in the nursery only requires comparing its $SO$
to the static boundaries of the nursery.

@@

Furthermore, we optimistically re-share pages during major collections.
This is another example where the STM-GC cooperation makes sense. We
regard private pages as additional memory that may be collected.
Pages should usually stay private for some time, but at some regular
intervals, we still want to optimistically re-share pages in case they actually
stay unmodified in the future. Major collections are allowed to do
that because all threads are synchronised, and we can re-share all
private pages that do not currently contain an object in some
transaction's  write set.

@@

The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.  For
each object, this code is called at most once between two minor collections.
First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!). Any reference we write to this object will
be kept alive and up-to-date during minor collections.

@@

We evaluate our system in a Python interpreter called
PyPy and compare to CPython.
\begin{description}
\item[PyPy] (version 2.3) is an implementation of an
  interpreter for the Python language. It has a special focus on speed
  and provides a just-in-time (JIT) compiler to speed up applications
  running on top of it. We compare between normal PyPy
  using a GIL and a PyPy with STM. All code is available at~\cite{pypy}.
\item[CPython] (version 2.7.6) is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\end{description}

@@

For the results in Figure~\ref{fig:scaling}, we
normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread ($12.3\%$ for all 4 threads together). Every thread
adds some overhead because during a commit, there is one more thread
which has to reach a safe point. Additionally, conflict detection
needs to check for conflicts in all concurrently running transactions.

@@

\subsection{Non-JIT Benchmarks}
First we run our benchmarks on three different interpreters: CPython
(GIL), PyPy with STM, and PyPy with the GIL (all without a JIT). The
results are shown in Figure~\ref{fig:performance-nojit}.

@@

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones ($avg=2.6\times$ speedup) and a little less for
the others ($avg=2.0\times$ speedup). The reason for this difference is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks to synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

@@

Since PyPy comes with a JIT~\cite{cfbolz09} to make its overhead compared to CPython
go away, we will now look at how well STM works together with it.

@@

We would like to regard enabling the JIT as a simple performance
enhancement, but that is not what happens in reality. First, since the
JIT~\cite{cfbolz09} bases its decision of what to compile on runtime
profiling, running in multiple threads, it may compile different
things in each run because of the non-deterministic thread-scheduling
of the operating system (OS). Second, it is able to perform some
advanced optimisations. Because compilation is already non-deterministic,
so are these optimisations. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.
For these reasons, the following results have to be taken with a grain
of salt.

@@

There are multiple attempts at removing the limits of HTM by means of
\emph{virtualising} HTM~\cite{rajwar05,chung06}. Indeed this may also
be the future direction of our research. So far, some of them depend
on additional hardware features. Or in the case of~\cite{chung06},
they have the assumption that most transactions fit the HTM capacity
and that virtualisation is only a fallback. Our use case, however,
features arbitrarily long transactions, which is why this assumption
first has to be verified again in this setting.

@@

% STM needs good performance:
The solution presented here is based on a software TM system.  As the
STM is a vital part of the interpreter, good performance is
essential. We present an interesting way to combine the existing
memory segmentation and virtual memory features of current CPUs to
lower the overhead of STM. This software implementation reduces
dramatically the cost of transactional memory and at the same time
provides an opportunity to integrate the STM system with garbage
collection.  This implementation leverages the processor's memory
segmentation hardware, which has lately not received much attention.
Unfortunately, processor architects seem to
undervalue this hardware resource -- in the evolution from the x86 to
the x86{-}64 architecture, 4 out of 6 segment registers have been
removed. This development potentially threatens the STM strategy
outlined here and we want to remind the architecture community that
this unit can support interesting implementations -- a viewpoint that
has also been expressed by other researchers (see also
\cite{bennet10}).

% results, outlook
The early results presented here are very encouraging. STM as a simple
GIL replacement scales well and yields an average speedup of $2.6\times$ for
embarrassingly parallel workloads on 4 threads.  When also used for
synchronisation in the form of atomic blocks, the average speedup
still reaches $2.0\times$.

To obtain an overall performance that is competitive with the
best-performing Python systems (Jython, PyPy), integration of the
STM-based approach with a JIT compiler is necessary. Once this
integration matures, the approach outlined here serves not only as a
simple GIL replacement but also provides a way forward towards
a parallel programming model for Python.

@@

\acks
We would like to thank the following people for their valuable
input and the many fruitful discussions: Carl F. Bolz, Michael Faes,
Pascal Felber, Maciej Fijalkowski, Patrick Marlier, and Aristeidis Mastoras.
